Lab 9: Multinomial and Ordered Probit Models

### 2023-01-04 Lab 8  BinomialModels/Lab8_BinomialModels_AnswerKey.Rmd
Add files via upload

---
title: 'Lab 7: Missing Data and Imputation:Answers'
author: "Eric Parajon"
header-includes:
  - \usepackage{enumerate}
  - \usepackage{graphicx}
output: pdf_document
---

Today's lab has been adapted from code by Chelsea Estancona, Isabel Laterzo, and Simon Hoellerbauer.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

@@

#Load in packages
pacman::p_load(tidyverse,
               rio,#for importing data
               arm)#inverse link function

@@

```{r}
colombia <- rio::import("Lab_8_data.dta")

@@

```{r}
ggplot(colombia, aes(x = as.factor(AUC))) +
  geom_bar() +
  labs(x = 'AUC Presence') +
  theme_bw()

@@

```{r}
#for a logit model, how should we specify family?

m1 <- glm(AUC ~ tribut_2_log + tribut_2_log_sq + pop_attacks + disbogota +
            drugs + pop_log, data=colombia, family = __________)

@@

```{r}
#for a probit model, how should we specify family?

m2 <- glm(AUC ~ ., data = colombia, family = ___________)
#what is the . doing above?
summary(m2)
```

@@

We might want to use our model to estimate predicted probabilities, as this is a readily-understood measure of how changes in our independent variable affect the probability of observing a certain outcome. In doing so, we'll want to think about a 'hypothetical case' for which we want to predict: for example, here, we'll calculate the predicted probability of the United Self-Defenses of Colombia or Autodefensas Unidas de Colombia (AUC) presence in a municipality with the mean tax income, mean attacks on population, mean distance from the capitol, drug crops, and the average population. *Note*: This is the so-called average case approach.

@@

#let's take a look!
invlogit #no parentheses from the arm package

@@

#Here use predict()
predict(_____, newdata = ________, type = 'response') #The type="response" option tells R to output probabilities of the form P(Y = 1|X)
```

@@

```{r}
## generate predicted probabilities for municipality with drugs and all other
## variables at their means
pred_output <- data.frame(Logit = predict(_____, newdata = _______, type = _______),
                          Probit = predict(____, newdata = ________, type = ______),
                          tax = new_data_2$tribut_2_log)

@@

Simulation strikes again! This time around, we will do similar things, but with non-Normal outcomes, where simulations are even more helpful.

Today, we're going to simulate predicted probabilities for a model with an interaction term and plot our results. We will use both the average case and the observed-case approaches.

@@

We're not going to worry a lot about confidence intervals today, but below is the code for them as well.

## CIs
pe <- apply(pp_sims, 2, mean)
lo <- apply(pp_sims, 2, quantile, prob = .025)
hi <- apply(pp_sims, 2, quantile, prob = .975)

@@

Generate an average predicted effect over the observed values of income for the first model. Plot this too.

#Load in packages
pacman::p_load(lmtest)

turnout<-import("turnout.csv")
#Fit the two models
  #m1
m1 <- glm(vote ~ race + educate + income, turnout, family =______)

  #m2
m2 <- glm(vote ~ race + educate + income + poly(age, 2, raw = T), turnout, 
         family =______))

#Model fit
lrtest(______)
AIC()
BIC()
AIC()
BIC()

# Average-Case Approach
#When we combine the average-case approach with simulation, we create a hypothetical design matrix, where we hold all continuous variables at their mean or another justifiable value, and then let the predictor of interest vary along some range.

sim_dat <- with(______, data.frame(intercept = 1, race = 1,
                                    educate = mean(educate), income = mean(income),
                                    age = seq(min(age), max(age), length.out = 100),
                                    age2 = (seq(min(age), max(age), length.out = 100))^2))

sim_betas <- mvrnorm(______)

@@

sim_gg <- data.frame(age = ______, pe = apply(sim_pp, 1, mean),
                     lo = ______,
                     hi = ______)

ggplot()

# Observed-Case Approach
#When taking the observed-case approach (see Hanmer and Kalkan (2013) for why this is the preferable approach) and using simulation to get confidence intervals, we have to do things slightly differently. Specifically, we will have to take an addition mean.

temporary <- model.matrix(m1)

#getting sim betas for m1
sim_betas_m1 <- mvrnorm(______)

#getting range of income
income_vec <- seq(from = ______,
                  length.out = 100)

#Here we use `map_dfc` function from `purrr` (part of the tidyverse) to we apply the function to each element in the vector, and then (because the function returns a vector) we get a dataframe in as output.

#This is doing a few things: 
#1. takes the temporary design matrix we created above, and then replaces *all* of the values of our predictor of interest with one value (taken from the hypothetical value vector).

#2. We then matrix multiply this new design matrix, with all actual values of our predictor of interest replaced by one hypothetical value of our predictor interest and with all the other values held at their observed values, by our simulated betas.

#3.  We then take the mean across the columns (`margins = 2`) because different columns represent different predicted probabilities, for each of the observations in our temporary design matrix, when we use different betas and when our predictor of interest has been replaced with one value.

#4. The call to `map_dfc` then returns a dataframe that has 1000 rows and 100 columns (1 column for each value of our hypothetical value vector).

pp <- map_dfc(income_vec, function(j) {
    temporary[, "income"] <- j

@@

#Pulling it all into a df and getting the mean of predicted values for each value of our hypothetical value vector, and then use the `quantile` with a `probs` argument equal to the bounds of our desired confidence interval (for example, if we want a 95% confidence interval, we will do `quantile(x, probs = c(0.025, 0.975))`)

@@

plotdat <- data.frame(plotdat, ______)

#We can then use these data to plot the mean predicted probability and our confidence bands.

#
colnames(______) <- c('pe', 'lo', 'hi', 'income')

ggplot()

>>>>

---
title: 'Lab 8: Binomiaal Models in R'
author: "TA: Eric Parajon"
output: pdf_document
---

Today's lab has been adapted from code by Chelsea Estancona, Isabel Laterzo, and Simon Hoellerbauer.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

#Load in packages
pacman::p_load(tidyverse,
               rio,
               arm)#for importing data
```

@@

```{r}
colombia <- rio::import("Lab_8_data.dta")
```

@@

```{r}
ggplot(colombia, aes(x = as.factor(AUC))) +
  geom_bar() +
  labs(x = 'AUC Presence') +
  theme_bw()
```

@@

```{r}
m1 <- glm(AUC ~ tribut_2_log + tribut_2_log_sq + pop_attacks + disbogota +
            drugs + pop_log, data=colombia, family = binomial(link = 'logit'))

@@

```{r}
m2 <- glm(AUC ~ ., data = colombia, family = binomial(link = 'probit'))
#what is the . doing above?
summary(m2)
```

@@

We might want to use our model to estimate predicted probabilities, as this is a readily-understood measure of how changes in our independent variable affect the probability of observing a certain outcome.

In doing so, we'll want to think about a 'hypothetical case' for which we want to predict: for example, here, we'll calculate the predicted probability of the United Self-Defenses of Colombia or Autodefensas Unidas de Colombia (AUC) presence in a municipality with the mean tax income, mean attacks on population, mean distance from the capitol, drug crops, and the average population.

*Note*: This is the so-called average case approach.

@@

#let's take a look!
arm::invlogit #no parentheses from the arm package

@@

predict(m1, newdata = new_data, type = 'response')
```

@@

```{r}
## generate predicted probabilities for municipality with drugs and all other
## variables at their means
pred_output <- data.frame(Logit = predict(m1, newdata = new_data_2, type = 'response'),
                          Probit = predict(m2, newdata = new_data_2, type = 'response'),
                          tax = new_data_2$tribut_2_log)

@@

We talked about simulations in Lab 4, when we talked about OLS. This time around, we will do similar things, but with non-Normal outcomes, where simulations are even more helpful!

Today, we're going to simulate predicted probabilities for a model with an interaction term, use the distribution of these simulated probabilities to define confidence intervals, and plot our results, using both the average case and the observed-case approaches.

@@

We're not going to worry about confidence intervals today, but below is the code for them as well.

```{r}
## CIs
pe <- apply(pp_sims, 2, mean)
lo <- apply(pp_sims, 2, quantile, prob = .025)
hi <- apply(pp_sims, 2, quantile, prob = .975)

@@

Generate an average predicted effect over the observed values of income for the first model. Plot this.

Required packages and data:
```{r, message=FALSE}
#Load in packages
pacman::p_load(lmtest)

turnout<-import("turnout.csv")

First, we fit the models:
```{r}
m1 <- glm(vote ~ race + educate + income, turnout, family = binomial(link = 'logit'))
m2 <- glm(vote ~ race + educate + income + poly(age, 2, raw = T), turnout, 
          family = binomial(link = 'logit'))

lrtest(m1, m2)
AIC(m1)
AIC(m2)
BIC(m1)
BIC(m2)
```

# Average-Case Approach

When we combine the average-case approach with simulation, we create a hypothetical design matrix, where we hold all continuous variables at their mean or another justifiable value, and then let the predictor of interest vary along some range.

```{r}
sim_dat <- with(turnout, data.frame(intercept = 1, race = 1,
                                    educate = mean(educate), income = mean(income),
                                    age = seq(min(age), max(age), length.out = 100),
                                    age2 = (seq(min(age), max(age), length.out = 100))^2))
```

In order to simulate different betas, we take a predetermined number of draws (1000 in this case here) from a multi-variate normal distribution centered around the coefficient estimates from our model, with a variance-covariance matrix also from our model. We then matrix multiply our hypothetical design matrix with our simulated betas. This means that we will have 1000 different predicted probabilities for each row in our hypothetical design matrix (in other words, we have 1000 different predicted probabilities for each element of the vector of values for our predictor of interest). We then have to make sure that we take the mean, and the quantiles we are interested in, along whichever dimension represents the elements of the hypothetical range vector. We can do this using `apply`. Note that in the example below, we are using `apply` along the rows of `sim_pp` because of how we constructed `sim_pp` (also using `apply`.) If we had not used `apply` and had instead simply done `as.matrix(sim_dat) %*% sim_betas`, the dimensions of `sim_pp` would have been reversed, and we would have to `apply` the mean and quantile functions along the columns, not the rows. 
```{r}
sim_betas <- mvrnorm(1e3, coef(m2), vcov(m2))

@@

sim_gg <- data.frame(age = sim_dat$age, pe = apply(sim_pp, 1, mean),
                     lo = apply(sim_pp, 1, quantile, probs = .025),
                     hi = apply(sim_pp, 1, quantile, probs = .975))
```

We can plot the predicted probabilities and our confidence bands.
```{r}
ggplot(sim_gg, aes(x = age, y = pe, ymin = lo, ymax = hi)) +
  geom_ribbon(aes(fill = '95% Confidence Interval'), alpha = .5) +
  geom_line(aes(color = 'Predicted Probability')) +
  labs(x = 'Age', y = 'Probability of Turnout') +
  theme_bw() +
  theme(legend.title = element_text(), legend.position = 'bottom',
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  scale_fill_manual('', values = 'dodgerblue4') +
  scale_color_manual('', values = 'grey30')
```

# Observed-Case Approach

When taking the observed-case approach (see Hanmer and Kalkan (2013) for why this is the preferable approach) and using simulation to get confidence intervals, we have to do things slightly differently. Specifically, we will have to take an addition mean. This is because whereas in the average-case approach, we had one row for each hypothetical value of our predictor of interest (however long we decided to make it), in the observed-case approach, we will have a whole matrix or dataframe for each hypothetical value of our predictor of interest.

We start by creating a temporary design matrix. This while start out as being the same exact design matrix used to fit the model, which is why we can just use `model.matrix()` on our model object.

```{r}
temporary <- model.matrix(m1)
```

We then have to make sure to once again simulate our betas, using our model object.
```{r}
#getting sim betas for m1
sim_betas_m1 <- mvrnorm(1e3, coef(m1), vcov(m1))
```

Just like with the average-case approach, we have to create the values we want our predictor of interest (*income* in this case) to take on. Just like in the average-case approach, these are really hypothetical values. Here, we will let income vary from its min to its max.
```{r}
#getting range of income
income_vec <- seq(from = min(turnout$income), to = max(turnout$income),
                  length.out = 100)
```

In the next step, we do several important things at once. We will do these things using the `map_dfc` function from `purrr` (part of the tidyverse). This works a lot like the `apply` family of functions, but is somewhat faster. It maps a function to a vector and then, because of the `_dfc` suffix, column binds the results of the function together. In other words, we apply the function to each element in the vector, and then (because the function returns a vector) we get a dataframe in as output. Note that we could do something similar with `lapply`, but then we would get a list of vectors.

```{r}
pp <- map_dfc(income_vec, function(j) {
    temporary[, "income"] <- j

@@

    return(pp)
    })
```

The function we apply to each element of our hypothetical predictor value vector does a few things:
1. It takes the temporary design matrix we created above, and then replaces *all* of the values of our predictor of interest with one value (taken from the hypothetical value vector).
This looks like the following:
```{r}
head(income_vec)

temporary[, "income"] <- income_vec[1]

head(temporary)
```
2. We then matrix multiply this new design matrix, with all actual values of our predictor of interest replaced by one hypothetical value of our predictor interest and with all the other values held at their observed values, by our simulated betas. This creates a matrix, which has as many rows as the rows of our data, and as many columns as there are samples of our betas.  
```{r}
dim(invlogit(temporary %*% t(sim_betas_m1)))
```
3. We then take the mean across the columns (`margins = 2`) because different columns represent different predicted probabilities, for each of the observations in our temporary design matrix, when we use different betas and when our predictor of interest has been replaced with one value. This means that we then have a mean predicted probability for each different set of betas when our predictor of interest is at a certain value.

The call to `map_dfc` then returns a dataframe that has 1000 rows and 100 columns (1 column for each value of our hypothetical value vector). Because we are interested in how the predicted probability changes according to the hypothetical values of our predictor of interest and because we currently have 1000 predicted probabilities for each of these hypothetical values (1 for each of our simulated betas), we once again use `apply` with `margins = 2`. We get the mean of predicted values for each value of our hypothetical value vector, and then use the `quantile` with a `probs` argument equal to the bounds of our desired confidence interval (for example, if we want a 95% confidence interval, we will do `quantile(x, probs = c(0.025, 0.975))`).

@@

plotdat <- data.frame(plotdat, income_vec)
```

We can then use these data to plot the mean predicted probability and our confidence bands.
```{r}
colnames(plotdat) <- c('pe', 'lo', 'hi', 'income')
head(plotdat)

ggplot(plotdat, aes(x = income, y = pe, ymin = lo, ymax = hi)) +
  geom_ribbon(aes(fill = '95% Confidence Interval (Simulation)'),
              alpha = .5) +
  geom_line(col = 'maroon') +
  labs(x = 'Income', y = 'Predicted Probability of Voting') +
  ylim(c(0, 1)) +
  scale_fill_manual('', values = 'darkorange') +
  theme_bw() +
  theme(legend.title = element_text(), legend.position = 'bottom',
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
```

### 2023-01-04 Lab 9  Multinom_orderedprob/Lab9_Multinom_OrderedProbit_Answers.Rmd
Add files via upload

>>>>

#Cleaning

hw<-hw %>% 
   mutate(warm = case_when(.$warm == 1 ~ "Strongly disagree", 
                          .$warm==2 ~ "Disagree", 
                          .$warm==3 ~ "Agree", 
                          .$warm==4 ~ "Strongly agree")) %>%
    mutate(warm = factor(warm, levels=c("Strongly disagree","Disagree","Agree","Strongly agree")))

#Create the model
warm_model <- polr(warm~male+white+age+ed+prst, 
           data = hw, method = "probit", Hess = TRUE)
summary(warm_model)

#What is the effect of increasing each predictor by one unit?
#How do the odds of moving from a lower to adjacent higher category change?
coefs <- warm_model$coefficients[1:5]
#odds ratios
exp(coefs)

#Check the proportional odds assumption for male
m1_prop_1 <- glm(I(as.numeric(warm) >= 2) ~ male, family = "binomial",
                 data = hw)

m1_prop_2 <- glm(I(as.numeric(warm) >= 3) ~ male, family = "binomial",
                 data = hw)

#male=0
check_1 <- m1_prop_1$coef[1] - m1_prop_2$coef[1]
check_1

#male=1
check_2 <- (m1_prop_1$coef[1] + m1_prop_1$coef[2]) - 
  (m1_prop_2$coef[1] + m1_prop_2$coef[2])
check_2

#Generate a cumulative predicted probability plot.
table(hw$male); table(hw$white); mean(hw$ed)
hw_2 <- expand.grid(male = 1, #modal value
                      white = 1, #modal value
                      ed=mean(hw$ed),
                      age=mean(hw$age),
                      prst = seq(from = min(hw$prst), to = max(hw$prst), length.out = 2293))

probs <- predict(warm_model, hw_2, type = "probs") #type = is different than for binomial model
head(rowSums(probs)) #just to double check
cumul_probs <- t(apply(probs, 1, cumsum)) #get the cumulative sum of the rows
#transpose into columns for plotting and merging with values of gpa
head(cumul_probs)
plot_data <- data.frame(hw$prst, cumul_probs)
ggplot(data=plot_data, aes(x=hw.prst))+
    geom_smooth(data=plot_data, aes(x=hw.prst, y=Strongly.disagree), se=FALSE)+
    geom_smooth(data=plot_data, aes(x=hw.prst, y=Disagree), se=FALSE)+
    geom_smooth(data=plot_data, aes(x=hw.prst, y=Agree), se=FALSE)+
      geom_smooth(data=plot_data, aes(x=hw.prst, y=Strongly.agree), se=FALSE)+
    geom_ribbon(data=plot_data, aes(ymin=0, ymax=Strongly.disagree, fill="Strongly.disagree"))+
    geom_ribbon(data=plot_data, aes(ymin=Strongly.disagree, ymax=Disagree,
                                    fill="Disagree"))+
    geom_ribbon(data=plot_data, aes(ymin=Disagree, ymax=Agree,
                                    fill="Agree"))+
         geom_ribbon(data=plot_data, aes(ymin=Agree, ymax=Strongly.agree,
                                    fill="Strongly.agree"))+
    labs(x="Values of prestige of occupation", y="Cumulative Predicted Probability")+
    scale_fill_discrete(name="Hard Work")+
    theme_bw()
